BOLD Parameter Estimation Using Particle Filters

### 2010-07-03 doc/thesis/methods.tex
some changes to methods

\section{Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
The primary design decision for resampling is the regularizing
kernel. As mentioned previously, the Gaussian kernel is extremely convenient,
because it is so easy to sample from. As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some amount of over-smoothing
won't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 10 to 20 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

The question of how often to resample is complex. The benefit of resampling
is that particles that were minimally important can give much more 
granularity in the middle of the distribution. At the same time, resampling
right after a batch of outliers can result in an incorrect distribution.
Because the memory of particle weights is not particularly long, "good" particles
could be removed by resampling very often. Additionally, because regularized
resampling is being with a large bandwidth, I did not want the particle filter
to be prevented from converging by resampling too often. Therefore, only
when the effective sample size dropped below 25 for two measurement points
in a row did I resample. This seemed to give good results, and avoided 
temporary drops in the $N_{eff}$ caused by outliers.

>>>>

\section{Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
The primary design decision for resampling is the regularizing
kernel. As mentioned previously, the Gaussian kernel is extremely convenient,
because it is very simple to sample from. As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some amount of over-smoothing
won't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 20 to 30 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

Because the regularized resample will almost certainly over-smooth,
it is necessary to wait until the $N_{eff}$ drops relatively low
(say below 25) before resampling is performed. As an additional 
measure against sharp drops in the $N_{eff}$ that may quickly raise
back up, resampling is only performed when two consecutive low
$N_{eff}$'s are found. 
The danger in waiting longer to resample is that if there were no
particles in the vicinity of the solution, then particle deprivation
can occur.  When this happens, the resampling
stage will have an inappropriately small variance which will lead to an 
inappropriately small support for the distribution. After this, the 
distribution will be unrecoverable. Thus, when the $N_{eff}$ gets down
to extremely low values, usually this is an indication of particle
deprivation; something that is also usually accompanied by extremely
fast drops in variance. When this happens, it is preferable 
to return to a previous, wider estimate of scale for the regularization
and continue from there. Thus to keep particle deprivation from 
affecting the results; a previous covariance matrix is 
used as the regularization bandwidth. In my algorithm I use
the last covariance matrix for which the $E_{eff}$ was greater than
the threshold. This guarantees that there was at least some amount of 
particle diversity, and also helps prevent converging prematurely.

When some sort of particle deprivation has not occurred, regularized
resampling with a large bandwidth will slow down convergence. Thus
as mentioned previously, to prevent over smoothing, the particle
filter algorithm used here waits until 
the effective sample size drops below 25 for two measurement points
in a row. This seemed to give good results, and avoids
temporary drops in the $N_{eff}$ caused by outliers.
